A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-02 tex/MercuryNumerical.tex
minor updates

\usepackage{subcaption} % subfigures

@@

We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
Said differently: a force $\vec F$ shows up by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity.
We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g. solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
}

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

@@

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun mercury system with relevant vectors.
		Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 10
\end{lstlisting}
Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
	#<...Code...>
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
\end{lstlisting}
\python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the finalization of the function.

Finally, we can express the evolution by a \texttt{while}-loop
\begin{lstlisting}%[float,floatplacement=H]
t     = 0
alpha = 0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
	t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha$ to zero in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrite the previous values and thus enter the next iteration.
The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
To describe at least one full orbital period, this time needs to be larger than $T_M$.
With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to include further objects of the \texttt{VPython} module \texttt{visual}.
Additional to the vector class, one has to include
\begin{lstlisting}
from vpython import vector, sphere, color, curve, rate
\end{lstlisting}
The class sphere will represent mercury and the sun in the simulation
\begin{lstlisting}
# define the initial coordinates; M = mercury, S = sun
M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
# and the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
# add a visible trajectory to mercury
M.trajectory = curve(color=color.white)
\end{lstlisting}
We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.

@@

If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}.
To get the perihelion motion, the term from equation (\ref{eq:newton_art}) has to be be added by choosing $\alpha=0$.
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories or analyze what happens when changing the time-steps $\Delta t$.
However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$.
Then the perihelion motion is better visible. This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^6$.

\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\caption{\label{fig:MercuryOrbit}%
		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
		All computations where executed with $\Delta t = 2 v_{M}(0) / c_a / 20$.
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
\begin{itemize}
\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits,
\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles
\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$
\item extrapolate $\delta \Theta (\alpha)$ to $\alpha=1$
\end{itemize}

There are several ways to find the position of the perihelion.
The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
This can be done within the $while$-loop.
To find identify a minimal distance to the sun, one has to know the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} in addition to the current position.
If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
The implementation can be realized as follows
\begin{lstlisting}
# set up vectors
vec_r_last = vec_rM0
turns      = 0
max_turns  = 10
list_perih = list()
# find perihelion for each turn and print it out
while turns < max_turns:
	vec_r_before_last = vec_r_last
	# store position of mercury in a new vector (since we will change M.pos)
	vec_r_last = vector(M.pos)
	#<...update mercury position...>
	# check if at perihelion
	if (vec_r_last.mag < M.pos.mag) and (vec_r_last.mag < vec_r_before_last.mag):
		list_perih.append(vec_r_last)
		turns = turns+1
\end{lstlisting}
To compute the angle between two vectors one uses the following
formula
 \begin{equation}
 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
	\, .
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# import function to resolve angle
from vpython import acos, pi, dot
# define function for angle extraction
def angle_between(v1, v2):
	return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
\end{lstlisting}
To account for the statistical errors (e.g., numerical rounding errors) one can be average over a few turns.
Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
\begin{lstlisting}
sum_angle=0.
for n in range(1, max_turns):
	# calculate angle
	sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
# display the average
print(sum_angle/(max_turns-1.))
\end{lstlisting}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tests of stability and error analysis}\label{sec:stability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Physical experiments in general, as do numerical simulations of physics, must always account for possible uncertainties.
It is important for the students to recognize this fact and to understand, that the cause of the uncertainties lies, e.g., in the finite time steps.
The students should therefore explore the source and size of the errors arising in this simulation at least qualitatively.

Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $\Delta t$ and $\alpha$.
As expected, the correlation between the number of turns and $\delta\Theta$ is linear and the slope depends solely on $\alpha$.
However, there is an offset between the lines for different $\Delta t$.
The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
This does not generate a problem for the procedure described in the previous section since the starting position is not considered. Still, it is helpful to be aware of this fact.
One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\todo[inline]{@all: where is this periodic oscillation coming from?}
	\caption{\label{fcc3}Precession growth of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
	One should pay special attention to the oscillations within the growth for relatively large time steps.
}
\end{figure}

To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
In figure \ref{fig: LargeDT} the mercury trajectory for $\Delta t=2 * v_M(0) / c_a \cdot 5$ is shown.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $\Delta t$.
Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/num-err.png}
	\caption{\label{fig: LargeDT}%
		Deviation of the trajectory from precise numeric result (black line; $\Delta t=2 v_M(0) / a_M / 20$) for large time steps (red line; $\Delta t=2 v_M(0) / a_M \times 2$).
		Both orbits are plotted for $T = 10 T_M$.
	}
\end{figure}

@@

Thanks to Einstein we indeed know that the conclusion formulated above is correct ---- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations. 
On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that there is probably some other dynamics going on that drives this difference.

>>>>

\usepackage{subcaption} % subfigures
\usepackage{caption}    % itemize in caption

@@

We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 17 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
Said differently: a force $\vec F$ shows up by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity.
We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
}.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{} (\url{https://www.python.org}) as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

@@

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun mercury system with relevant vectors.
		Because mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_M$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
	# Compute the strength of the acceleration
	aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
\end{lstlisting}
\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.

Finally, we can describe the evolution by a \texttt{while}-loop
\begin{lstlisting}%[float,floatplacement=H]
t     = 0
alpha = 0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
	t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha=0$ in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrites the previous values and thus enter the next iteration.
The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
To describe at least one full orbital period, this time needs to be larger than $T_M$.
With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
Note that the exact time it takes for mercury complete a full revolution depends on the initial coordinates and velocities as well as the accuracy of the computation.
One might encourage students to analyze this in the beginning.

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to \texttt{import} further objects from the \texttt{VPython} module.
Additional to the vector class, one has to include
\begin{lstlisting}
from vpython import vector, sphere, color, curve, rate
\end{lstlisting}
The class sphere will represent mercury and the sun in the simulation
\begin{lstlisting}
# define the initial coordinates; M = mercury, S = sun
M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
# and the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
# add a visible trajectory to mercury
M.trajectory = curve(color=color.white)
\end{lstlisting}
We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.

@@

If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe its effects on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in equation (\ref{eq:newton_art}) has to be "turned on" for $\alpha\neq0$ [figure \ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
Then the perihelion motion is better visible.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different mercury orbits for:
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (for modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
	}
\end{figure}

@@

To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
\begin{itemize}
\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits;
\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles;
\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$;
\item interpolate $\delta \Theta (\alpha)$ to $\alpha=1$ using that $\delta \Theta (0) = 0$ (modulo numerical errors).
\end{itemize}

There are several ways to find the position of the perihelion.
The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
This can be done within the $while$-loop.
To find identify a minimal distance to the sun, one has to know the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} in addition to the current position.
If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
The implementation can be realized as follows\footnote{%
	Note that one stores \code{vec\_r\_last} by making a copy of \code{M.pos}: \code{vec\_r\_last = vector(M.pos)}.
	This is essential because otherwise, if one changes \code{M.pos}, one would automatically change \code{vec\_r\_last} as well.
}
\begin{lstlisting}
# set up vectors
vec_r_last = vec_rM0
turns      = 0
max_turns  = 10
list_perih = list()
# find perihelion for each turn and print it out
while turns < max_turns:
	vec_r_before_last = vec_r_last
	# store position of mercury in a new vector (since we will change M.pos)
	vec_r_last = vector(M.pos)
	#<...update mercury position...>
	# check if at perihelion
	if (vec_r_last.mag < M.pos.mag) and (vec_r_last.mag < vec_r_before_last.mag):
		list_perih.append(vec_r_last)
		turns = turns+1
\end{lstlisting}
To compute the angle between two vectors one uses the following
formula
 \begin{equation}
 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
	\, .
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# import function to resolve angle
from vpython import acos, pi, dot
# define function for angle extraction
def angle_between(v1, v2):
	return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
\end{lstlisting}
To account for the statistical errors (e.g., numerical rounding errors) one can average over a few turns.
Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
\begin{lstlisting}
sum_angle=0.
for n in range(1, max_turns):
	# calculate angle
	sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
# display the average
print(sum_angle/(max_turns-1.))
\end{lstlisting}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tests of stability and error analysis}\label{sec:stability}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{@all: where is this periodic oscillation coming from?}
Physical experiments in general, as do numerical simulations of physics, must always account for possible uncertainties.
It is important for the students to recognize this fact and to understand, that the cause of the uncertainties lies, e.g., in the finite time steps.
The students should therefore explore the source and size of the errors arising in this simulation at least qualitatively.

Consider figure \ref{fcc3}, where the angle of the perihelion growth $\delta \Theta$ is shown for the first $10$ turns for different choices of $\Delta t$ (colors) and $\alpha$ (columns).
As expected, the correlation between the number of turns and $\delta\Theta$, for sufficiently small time-steps $\Delta t$, is constant on average and the values depend solely on $\alpha$.
However, there is an offset between the lines for different $\Delta t$.
The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
This does not generate a problem for the procedure described in the previous section since the starting position is not considered. Still, it is helpful to be aware of this fact.
One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\caption{\label{fcc3}Precession growth of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
	One should pay special attention to the oscillations within the growth for relatively large time steps.
}
\end{figure}

To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
In figure \ref{fig:MercuryOrbit-a0-small-dt-large} the mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$ is shown.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $\Delta t$.
Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.

@@

Thanks to Einstein we indeed know that the conclusion formulated above is correct --- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations. 
On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that there is probably some other dynamics going on that drives this difference.
